---
title: "eagle algorithm"
output: html_notebook
---

## Eagle Algorithm Simple Version

Suppose that there is no repeated observation (Z = I), no missing value in genotype data.\
And the phenotype data is of single dimension.

The name of functions follow such style : do.something(), do.some.vector(), do.some.Matrix()\
The name of variables follow such style : some_data, some_vector, Some_Matrix

```{r}
require('MM4LMM')
require("glmnet")
require("testthat")

do_test <- TRUE
use_data1 <- TRUE
use_data2 <- FALSE
```

#### Load data if do test

```{r}
if (do_test) {
  if (use_data1){
    data(QTLDetectionExample)
    Geno <- QTLDetectionExample$Genotype
    Pheo <- QTLDetectionExample$Phenotype
    Kinship <- QTLDetectionExample$Kinship
    message("Example data loaded.")
  } else if (use_data2) {
    DIR <- "https://raw.githubusercontent.com/geo047/Example_Data/master/"
    if (!file.exists("mapex.dat")) download.file(paste0(DIR, "mapex.dat"))
    if (!file.exists("phenoex.dat")) download.file(paste0(DIR, "phenoex.dat"))
    if (!file.exists("genoex.dat.zip")) download.file(paste0(DIR, "genoex.dat.zip"))
    if (!file.exists("genoex.dat")) unzip("genoex.dat.zip")
  }
}
```

#### Syntactic Sugar

```{r}
# inverse of a matrix M
inv <- function(M) chol2inv(chol(M))
# faster multiple cross product of a list of matrix Mlist
mcp <- function(...) Reduce(function(u,v) crossprod(t(u),v), list(...))
# absolute value of the self-adjoint operator of a random normal distribution matrix plus a diagonal matrix
arndm <- function(nrow) abs(crossprod(matrix(data = rnorm(nrow^2), nrow = nrow))) + diag(nrow = nrow)
```

```{r}
if (do_test) {
  nrow = 9
  M1 <- arndm(nrow)
  M2 <- arndm(nrow)
  M3 <- arndm(nrow)
  M4 <- arndm(nrow)
  
  res1 <- M1 %*% M2 %*% M3 %*% M4
  res2 <- mcp(M1,M2,M3,M4)
  
  test_that("Test of Syntactic Sugar", {
    expect_lt(max(abs(inv(M1) %*% M1 - diag(nrow = nrow))), 1e-10)
    expect_lt(max(abs(M1 %*% inv(M1) - diag(nrow = nrow))), 1e-10)
    expect_lt(max(abs(res1 - res2)), 1e-10)
  })
  
  plot(res1, res2, main="compare the result of cross product of a list of matrix", 
       xlab="cross product by %*%", ylab="cross product by mcp", col="blue")
}
```


#### Build X matrix, X is the selected locus genotype data

dimension of X : n x p\
p = length(selected_locus)\
n is the number of observations

```{r}
build.X <- function (geno, selected_locus = NULL) {

  # dim of geno : n x p, n is the number of observation, p is the number of SNPs
  X <- matrix(data = 1, nrow = nrow(geno), ncol = 1)
  colnames(X) <- c("intercept")
  
  if (!is.null(selected_locus)) {
    X <- cbind(X, geno[, selected_locus])
  }
  
  return(as.matrix(X))
}
```

#### test cases of the funtion build.X

```{r}
if (do_test) {

  X <- build.X(geno = Geno)
  
  test_that("Testing build.X result dimension and value", {
    expect_equal(dim(X), c(dim(Geno)[1],1))
    expect_equal(dim(build.X(geno = Geno, 1)), c(dim(Geno)[1],2))
    expect_equal(dim(build.X(geno = Geno, c(1,2))), c(dim(Geno)[1],3))
    expect_equal(length(unique(X)), 1)
    expect_equal(unique(X)[1], 1)
  })
  
  X <- build.X(geno = Geno, c(1,2,3))
  
  message("[PASS] test cases of build.X\n")
  
  head(X)
}
```

#### build M_s matrix, M_s is the unselected locus genotype data

```{r}
build.M_s <- function(geno, selected_locus = NULL) {

  if (is.null(selected_locus)) {
    return(geno)
  }
  
  return(geno[, - selected_locus])
}
```

#### Test case of the function build.M_s

```{r}
if (do_test) {
  M_s0 <- build.M_s(geno = Geno, selected_locus = NULL)
  M_s  <- build.M_s(geno = Geno, selected_locus = c(1,2,3))

  test_that("Testing build.M_s result dimension and value", {
    expect_equal(dim(M_s0), dim(Geno))
    expect_equal(dim(M_s), c(dim(Geno)[1],dim(Geno)[2]-3))
  })
  
  message("[PASS] test cases of build.M_s\n")
  
  head(M_s)
}
```

#### build Kinship Matrix

```{r}
build.Kinship <- function(M_s) {
  L <- ncol(M_s) 
  # standarize M_s by columns
  M_s <- scale(M_s)
  MMt <- tcrossprod(M_s) / L
  return(MMt)
}
```

#### test cases of the funtion build.Kinship

```{r}
if (do_test) {
  M_s  <- build.M_s(geno = Geno, selected_locus = c(1,2,3))
  K <- build.Kinship(M_s = M_s)

  test_that("Test build.Kinship result dimension", {
    expect_equal(dim(K), c(dim(M_s)[1],dim(M_s)[1]))
  })
  
  message("[PASS] test cases of build.Kinship\n")
  
  print(sprintf("dim(Kinship): %d x %d", dim(K)[1],dim(K)[2]))
}
```

#### Calculate the parameters of Models $\hat{\mu}, \hat{\sigma_G}^2, \hat{\sigma_E}^2$

```{r}
calc.variances <- function (Y, X, Kinship, init = NULL, Method = "Reml") {
  # X : fixed effect
  
  n <- length(Y)
  
  ## Build the VarList object
  VL <- list(Additive = Kinship , Error = diag(1,n))
  
  ## Perform inference by MM4LMMm, todo : save the old sigma
  res <- MMEst(Y = Y, Cofactor = X, VarList = VL, Init = init, Method = Method)
  
  v <- list(vg = res$NullModel$Sigma2["Additive"],
            ve = res$NullModel$Sigma2["Error"],
            Beta = res$NullModel$Beta)
  
  if (Method == "Reml") {
    v$loglik <- res$NullModel$`LogLik (Reml)`
    v$method <- Method
  } else if (Method == "ML") {
    v$loglik <- res$NullModel$`LogLik (ML)`
    v$method <- Method
  }
  
  return(v)
}
```

#### Test case of the function calc.variances

```{r}
if (do_test) {
  
  s <- c(1,2)
  X <- build.X(geno = Geno, selected_locus = s)
  M_s <- build.M_s(geno = Geno, selected_locus = s)
  K <- build.Kinship(M_s = M_s)
  res <- calc.variances(Y = Pheo, X = X, Kinship = K, Method = "ML")
  
  test_that("Test calc.variances result", {
    expect_length(res$Beta, length(s)+1)
    expect_true(is.numeric(res$vg))
    expect_true(is.numeric(res$ve))
  })
  
  message("[PASS] test cases of calc.variances\n")
  
  print(res)
}
```

#### Calculate $\hat{\beta}, var(\hat{\beta})$

```{r}
calc.beta <- function(Y, M_s, Kinship, mu, v) {
  # M_s : genetic effect
  calc.diag_var_beta1 <- function(D, V, lambda, vg, L) vg * diag(mcp(V, diag((D**2)/(D**2+lambda*L)), t(V))) / L
  calc.diag_var_beta2 <- function(D, V, lambda, vg, L) vg*crossprod(t(V*V), (D**2)/(D**2+lambda*L))/L
  
  lambda <- v$ve / v$vg
  X_s <- scale(M_s)
  L <- length(colnames(X_s))
  K_ <- Kinship + diag(x = lambda, nrow = nrow(Kinship))
  inv_K_ <- inv(K_)
  SVD <- svd(X_s)
  beta <- mcp(t(X_s), inv_K_, (Y - mu)) / L
  diag_var_beta1 <- calc.diag_var_beta1(D = SVD$d, V = SVD$v, lambda = lambda, vg = v$vg, L = L)
  diag_var_beta2 <- calc.diag_var_beta2(D = SVD$d, V = SVD$v, lambda = lambda, vg = v$vg, L = L)
  return(list(beta = beta, diag_var_beta1 = diag_var_beta1, diag_var_beta2 = diag_var_beta2))
}
```

#### Test case of the function calc.beta

```{r}
if (do_test) {
  s <- c(1,2)
  X <- build.X(geno = Geno, selected_locus = s)
  M_s <- build.M_s(geno = Geno, selected_locus = s)
  K <- build.Kinship(M_s = M_s)
  vars <- calc.variances(Y = Pheo, X = X, Kinship = K)
  mu <- X %*% vars$Beta
  beta_hat <- calc.beta(Y = Pheo, M_s = M_s, Kinship = K, mu = mu, v = vars)
  
  # compare diag_var_beta
  X_s <- scale(M_s)
  L <- length(colnames(X_s))
  lambda <- vars$ve / vars$vg
  K_ <- K + diag(lambda * L, nrow = nrow(K))
  inv_K_ <- inv(K_)
  diag_var_beta <- diag(vars$vg * mcp(t(X_s), inv_K_, X_s) / L)
  
  # compare beta
  fit <- glmnet(x = M_s, y = Pheo - mu, alpha = 0, lambda = lambda)
  
  test_that("calc.beta result", {
    expect_equal(length(beta_hat$beta), ncol(M_s))
    expect_lt(max(abs(beta_hat$diag_var_beta1 - diag_var_beta)), 1e-13)
    expect_lt(max(abs(beta_hat$diag_var_beta2 - diag_var_beta)), 1e-13)
  })
  
  message("[PASS] test cases of calc.beta")
  
  plot(beta_hat$diag_var_beta1, diag_var_beta)
  plot(beta_hat$diag_var_beta2, diag_var_beta)
  plot(fit$beta, beta_hat$beta)
  
  print(beta_hat)
}
```

#### find QTL - Quantitative trait locus

```{r}

calc.snps_scores <- function(beta, var_beta) beta/var_beta

find.QTL <- function (scores, qtl_names) {
  
  max_idx <- which.max(scores)
  max_name <- rownames(scores)[max_idx]
  
  return(list(score = scores[max_idx], 
              name = max_name,
              idx = which(qtl_names == max_name)))
}

```

#### Test case of the function calc.snps_scores

```{r}
if (do_test) {
  s <- c(1,2)
  X <- build.X(geno = Geno, selected_locus = s)
  M_s <- build.M_s(geno = Geno, selected_locus = s)
  K <- build.Kinship(M_s = M_s)
  vars <- calc.variances(Y = Pheo, X = X, Kinship = K)
  mu <- tcrossprod(X, t(vars$Beta))
  beta_hat <- calc.beta(Y = Pheo, M_s = M_s, Kinship = K, mu = mu, v = vars)
  scores <- calc.snps_scores(beta = beta_hat$beta, var_beta = beta_hat$diag_var_beta1)
  res <- find.QTL(scores = scores, colnames(Geno))
  
  test_that ("calc.snps_scores result", {
    expect_equal(dim(scores), c(dim(M_s)[2],1))
    expect_equal(res$name, colnames(Geno)[res$idx])
  }) 
  
  message("[PASS] test cases of calc.beta")
  
  print(res)
}
```

#### calculate the score of a new model

```{r}
calc.model_score <- function(loglik, nobs, l, s, gamma) {
  # l : total number of snps
  # s : number of selected locus
  # loglik: Log Likelihood
  # gamma: extended BIC parameter
  
  calc.BIC    <- function(nobs, loglik, l) -2 * loglik + l * log(nobs)
  calc.extBIC <- function(bic, l, s, gamma) bic + 2 * gamma * lchoose(l, s)
  
  bic <- calc.BIC(nobs, loglik, l)
  xbic <- calc.extBIC(bic, l, s, gamma)
  
  return(xbic)
}
```

#### Test case of the function calc.model_score

```{r}
if (do_test) {
  score <- calc.model_score(loglik = 10, nobs = 10, l = 100, s = 10, gamma = 0.6)
  print(score)
}
```


### Algorithm: EAGLE, cas: Z = I

```{r}

simpleEagle <- function (pheno_mat, geno_mat, gamma = 0.6, max_itr_num = 100, fixit = FALSE) {
  
  #### init loop 
  total_SNPs_num <- ncol(geno_mat)
  continue <- TRUE
  itr_num <- 1
  selected_locus <- c()
  selected_locus_names <- c()
  model_score_history <- c()
  vars_param_history <- list()
  beta_hat_param_history <- list()
  
  while (continue) {
    
    print(sprintf("Iteration: %d ", itr_num))
    
    #### #### #### ####    STEP 1 model building    #### #### #### ####
    # mixed model: Y = X%*%tou + Z %*% Ug + e
    X <- build.X(geno = geno_mat, selected_locus = selected_locus)
    
    #### build Kinship Matrix
    M_s <- build.M_s(geno = geno_mat, selected_locus = selected_locus)
    K <- build.Kinship(M_s)
    
    #### calculate expectation and variances by MM4LMM
    vars <- calc.variances(Y = pheno_mat, X = X, Kinship = K)
    beta_hat <- calc.beta(Y = pheno_mat, M_s = M_s, Kinship = K, 
                          mu = tcrossprod(X,t(vars$Beta)), v = vars)
    
    #### #### #### ####    STEP 2 model evaluation   #### #### #### #### 
    # calculate the score of the model 
    model_score <- calc.model_score(loglik = vars$loglik, nobs = nrow(M_s), 
                                    l = total_SNPs_num, s = length(selected_locus), 
                                    gamma = gamma) 
    
    print(sprintf("score of the model: %f ", model_score))
    
    #### Termination Conditions: 
    # 1. new score of model < last one 
    # Question: Extend BIC is going up or down while the model is being promoted ? 
    if (!fixit && !is.null(model_score_history)) {
      if (model_score < max(model_score_history)) {
        itr_num <- itr_num - 1
        selected_locus <- selected_locus[-itr_num]
        selected_locus_names <- selected_locus_names[-itr_num]
        print("[Termination Message] The score of the new model cannot beat the previous one.")
        break
      }
    }
    
    model_score_history <- c(model_score_history, model_score)
    vars_param_history[[itr_num]] <- vars
    beta_hat_param_history[[itr_num]] <- beta_hat$beta
    
    # 2. iteration number > maximum iteration number
    if (itr_num > max_itr_num ) {
      print("[Termination Message] The iteration number reach the max.")
      break
    }
    
    # 3. all SNPs are going to be selected
    # To avoid pass abnormal values to MM4LMM
    if (itr_num >= total_SNPs_num - 1) {
      print("[Termination Message] Too many locus are selected, it seems wired. ")
      break
    }
    
    #### #### #### ####    STEP 3 model selection   #### #### #### #### 
    scores <- calc.snps_scores(beta = beta_hat$beta, var_beta = beta_hat$diag_var_beta1)
    new_selected_loci <- find.QTL(scores = scores, colnames(geno_mat))
    selected_locus <- c(selected_locus, new_selected_loci$idx)
    selected_locus_names <- c(selected_locus_names, new_selected_loci$name)
    
    print(sprintf("new selected loci: %s ", new_selected_loci$name))
    
    itr_num <- itr_num + 1
  }
  
  return(list(SNPs_idx = selected_locus, 
              SNPs_names = selected_locus_names, 
              fixed_effect_param = vars$Beta,
              var_additive = vars$vg,
              var_error = vars$ve,
              genetic_effect_param = beta_hat_param_history[[itr_num]],
              model_score = model_score_history[[itr_num]], 
              model_score_history = model_score_history))
}
```

#### test case of the algorithm

```{r}
if (do_test) {
  if (use_data1){
    data(QTLDetectionExample)
    Geno <- QTLDetectionExample$Genotype
    Pheo <- QTLDetectionExample$Phenotype
    res <- simpleEagle(pheno_mat = Pheo, geno_mat = Geno)
    
    test_that("EAGLE, Z = I", {
      expect_equal(length(res$SNPs_idx), length(res$SNPs_names))
      expect_equal(length(res$fixed_effect_param), length(res$model_score_history))
      expect_equal(length(res$SNPs_idx) + 1, length(res$fixed_effect_param))
    })
    
    message("[PASS] test cases of EAGLE")
    
    print(res)
  } 
}

```

